---
title: "Homework 10"
author: "Solutions"
output: html_document
---

```{css, echo = F}
.rsoln {
background-color: #E6E6FA;
}

.soln {
color: #5858d0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE,
                      fig.height = 3)
library(tidyverse)
```

```{r, echo=F}
old <- theme_get()

theme_set(theme(panel.background = element_rect(fill = "#ededfb")))
```


# Preliminaries

- This file should be in `STAT240/homework/hw10` on your local computer.
- Download `parental_leave.csv` and `happiness_2019` to `STAT240/data`.


# Problem 1

Two sets of data were used in the equal-variances T hypothesis test of

$$H_0: \mu_1 - \mu_2 = 2 \quad \text{versus} \quad H_A: \mu_1 - \mu_2 < 2$$

We know that $s_1 = 2, s_2 = 3, n_1 = 16$, and $n_2 = 15$. If the p-value is 0.001298716, what is the observed difference in means, $\bar{x}_1 - \bar{x}_2$?

[First, we should figure out what the test statistic was. The one-sided p-value is 0.001298716, which means that the test statistic must have area 0.001298716 below it in the lower tail of a t with $16 + 15 − 2 = 29$ degrees of freedom. We can use `qt()` to find that specific value.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
qt(0.001298716, df = 29)
```

[The observed test statistic is $-3.29542$.  We set this equal to the T test statistic formula, then solve for $\bar{x}_1 - \bar{x}_2$.  First, we need to calculate a pooled standard deviation.  This is given by 
$$s_p \;=\; \sqrt{\frac{(16-1)2^2 + (15-1)3^2}{16 + 15 - 2}} \;=\; 2.533.$$
So, we have
$$t_{obs} \;=\; \frac{\bar{x}_1 - \bar{x}_2 - 2}{2.533\sqrt{\frac{1}{16} + \frac{1}{15}}} \;=\; -3.29542.$$
Solving for $\bar{x}_1 - \bar{x}_2$ shows that the observed difference is -1.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
diff <- -3.29542*(2.533*sqrt(1/16 + 1/15)) + 2

diff
```


# Problem 2

Continue working with the parental leave data from homework 9.

```{r}
leave <- read_csv("../../data/parental_leave.csv")

# Remove spaces from column names
names(leave) <- make.names(names(leave))

# Tidy numeric columns
leave <- leave %>%
  mutate_at(vars(Paid.Maternity.Leave, Unpaid.Maternity.Leave,
                 Paid.Paternity.Leave, Unpaid.Paternity.Leave),
            as.numeric) %>%
  replace(is.na(.), 0)
```

(a) Re-do the basic data tidying from homework 9 problem 5(a): Mutate a column for total paid leave by adding the weeks of paid maternity and paternity leave. Optionally, give the columns shorter names that are easier to reference.

```{r, class.source="rsoln", class.output="rsoln"}
leave <- leave %>%
  rename(mat_paid = Paid.Maternity.Leave,
         pat_paid = Paid.Paternity.Leave,
         mat_unpaid = Unpaid.Maternity.Leave,
         pat_unpaid = Unpaid.Paternity.Leave) %>%
  mutate(total_paid = mat_paid + pat_paid)
```

(b) Consider comparing companies in "Educational Services: College & Universities" (ESCU) to companies in "Healthcare: Hospitals & Clinics" (HHC).  Create and print a table summarizing the mean, standard deviation, and sample size for both groups.

```{r, class.source="rsoln", class.output="rsoln"}
escu_vs_hhc <- leave %>% 
  filter(Industry %in% c("Educational Services: College & Universities",
                         "Healthcare: Hospitals & Clinics")) %>% 
  group_by(Industry) %>% 
  summarize(avg_leave = mean(total_paid), sd_leave = sd(total_paid), n = n())

xbar1 <- escu_vs_hhc$avg_leave[1]
xbar2 <- escu_vs_hhc$avg_leave[2]
s1 <- escu_vs_hhc$sd_leave[1]
s2 <- escu_vs_hhc$sd_leave[2]
n1 <- escu_vs_hhc$n[1]
n2 <- escu_vs_hhc$n[2]
```

(c) Assess the equal variances assumption for these two groups of data by comparing the ratio of standard deviations.  Even if the standard deviations are similar "enough", why might we want to not assume equal variances?

[To safely assume equal variances, we want the sample SDs to be within a factor of two, which is technically true for this set of data.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
s1 / s2
```

[In any case we are not required to assume equal variances, since the Welch T test is more general than the equal variances T test.  We might want to play it safe and not assume the variances are the same.  If the variances are not actually equal, then the equal-variances test might give us inaccurate results.]{.soln}

# Problem 3

Assuming unequal variances, build a 95% CI for the difference in average total weeks of paid leave between the ESCU and HHC industries and interpret the results. You must do this by hand, but you can check your results with `t.test()`.

```{r, class.source="rsoln", class.output="rsoln"}
escu_vs_hhc <- leave %>% 
  filter(Industry %in% c("Educational Services: College & Universities",
                         "Healthcare: Hospitals & Clinics")) %>% 
  group_by(Industry) %>% 
  summarize(avg_leave = mean(total_paid), sd_leave = sd(total_paid), n = n())

xbar1 <- escu_vs_hhc$avg_leave[1]
xbar2 <- escu_vs_hhc$avg_leave[2]
s1 <- escu_vs_hhc$sd_leave[1]
s2 <- escu_vs_hhc$sd_leave[2]
n1 <- escu_vs_hhc$n[1]
n2 <- escu_vs_hhc$n[2]
```

[After gathering data summaries, we need to calculate the approximate degrees of freedom to use for our T reference distribution.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
w_numer <- (s1^2/n1 + s2^2/n2)^2
w_denom <- (s1^4/(n1^2*(n1-1)) + s2^4/(n2^2*(n2-1)))

w <- w_numer / w_denom
```

[Now, we can find all of the components of our CI.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
pt_est <- xbar1 - xbar2
se <- sqrt(s1^2/n1 + s2^2/n2)

cv <- qt(0.975, df = w)

c(pt_est - cv*se, pt_est + cv*se)
```

[We are 95% confident that the true difference between average ESCU and HHC paid leave is within (1.509, 6.7).]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
obs_escu <- leave %>%
  filter(Industry == "Educational Services: College & Universities") %>%
  pull(total_paid)

obs_hhc <- leave %>%
  filter(Industry == "Healthcare: Hospitals & Clinics") %>%
  pull(total_paid)

t.test(obs_escu, obs_hhc)
```

# Problem 4

Does the educational industry give more paid leave than the healthcare industry?  Perform a hypothesis test of 
$$H_0: \mu_{ESCU} - \mu_{HHC} = 0 \quad \text{versus}\quad H_A: \mu_{ESCU} - \mu_{HHC} > 0$$
with $\alpha = 0.05$, assuming unequal variances.  You must do this by hand, but you can check your results with `t.test()`.

[Our test statistic is based on the difference in the observed means, and the null distribution is a T with $w$ degrees of freedom (found in problem 1). Our p-value is the area above our test statistic.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
se <- sqrt(s1^2/n1 + s2^2/n2)

test_stat <- ((xbar1 - xbar2) - 0)/se

test_stat

pt(test_stat, df = w, lower.tail = F)
```

[We get a very small p-value of 0.001, which is less than $\alpha = 0.05$.  We have evidence that the educational industry gives more paid leave than the healthcare industry, on average.  We get the same result with `t.test()`.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
t.test(obs_escu, obs_hhc, alternative = "greater")
```


# Problem 5

Repeat the by-hand test in problem 4, but instead write hypotheses for $\mu_{HHC} - \mu_{ESCU}$ (the order of subtraction is switched).  Show that you get identical results to problem 4.

[When we reverse the order of subtraction, we need to reverse the direction of the alternative hypothesis.  If we expect ESCU to be the bigger group, then $\mu_{HHC} - \mu_{ESCU}$ should be negative.  We have:
$$H_0: \mu_{HHC} - \mu_{ESCU} = 0 \quad \text{versus} \quad H_A: \mu_{HHC} - \mu_{ESCU} < 0$$]{.soln}

[Our test statistic has the reverse order of subtraction, which gives us the same test statistic as in problem 2, but with the opposite sign.  Now, we calculate a p-value by looking at the area below our test statistic.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
test_stat <- ((xbar2 - xbar1) - 0)/se

test_stat

pt(test_stat, df = w)
```

[We get an identical p-value of 0.001.]{.soln}


# Problem 6

Now, consider the "Technology: Software" industry.  Perform the appropriate hypothesis test to determine whether there is a difference in weeks of paid maternity and paternity leave within this industry.  You must do this by hand, but you can check your results with `t.test()`.

[Because maternity and paternity leave are both decided within a company, we have paired data, and we should perform our analysis at the "company level".  First, calculate the difference in maternity - paternity paid leave for each software company.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
leave_differences <- leave %>%
  filter(Industry == "Technology: Software") %>%
  mutate(diffs = mat_paid - pat_paid) %>%
  pull(diffs)
```

[Now, we need to perform a one-sample T test.  Our hypotheses are
$$H_0: \mu_{diff} = 0 \quad \text{versus} \quad H_A: \mu_{diff} \neq 0$$]{.soln}

[Let's gather the summaries of the differences and calculate a test statistic and two-sided p-value.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
xbar <- mean(leave_differences)
s <- sd(leave_differences)
n <- length(leave_differences)

test_stat <- (xbar - 0) / (s / sqrt(n))

test_stat

2*pt(test_stat, df = n-1, lower.tail = F)
```

[We have an extremely small p-value, which corresponds to very strong evidence against the null.  We have evidence of a difference in the amount of paid maternity and paternity leave offered in the software industry.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
t.test(leave_differences)
```


# Problem 7

Match the six values of correlation to the scatterplots in `p7_choices.png`. Briefly justify your choices.

- $r = -0.85$

[Graph C shows $r = -0.85$. There is a strong, but not perfect, negative correlation.]{.soln}

- $r = -0.74$

[Graph D shows $r = -0.74$. There a negative correlation that is not quite as strong as shown in graph C.]{.soln}

- $r = 0.08$

[Graph B shows $r = 0.08$.  There is no clear linear relationship between x and y, so correlation is approximately 0.]{.soln}

- $r = 0.44$

[Graph F shows $r = 0.44$.  There is a visible positive trend between x and y, but it is not very strong.]{.soln}

- $r = 0.98$

[Graph A shows $r = 0.98$.  There is a very strong positive relationship between x and y, but there is still a bit of random scatter present.]{.soln}

- $r = 1$

[Graph E shows $r = 1$.  There is a perfect linear relationship with no random error.]{.soln}


# Problem 8

The [Happiness index](https://www.kaggle.com/datasets/sougatapramanick/happiness-index-2018-2019) data contains information on the happiness index of different countries from 2018-2019.  We will focus on the happiness index (`Score`) and GDP per capita of countries in 2019.

```{r}
happiness <- read_csv("../../data/happiness_2019.csv")

# Remove spaces from column names
names(happiness) <- make.names(names(happiness))
```

(a) Create a scatterplot of GDP per capita (x) versus happiness index (y) and calculate the correlation between length and weight.  Comment on the strength and magnitude of the linear relationship, and whether a linear model would be appropriate for this data.

```{r, class.source="rsoln", class.output="rsoln"}
happiness %>%
  ggplot(aes(x = GDP.per.capita, y = Score)) +
  geom_point()

x <- happiness$GDP.per.capita
y <- happiness$Score

cor(x, y)
```

[Both the scatterplot and correlation indicate a fairly strong positive linear relationship.  Visually, the points are scattered around a line, so a linear model seems like a good fit for this data.]{.soln}

(b) Calculate the slope and intercept of a least-squares linear regression model for GDP per capita versus happiness index.  Do this with the "by hand formulas" and check your work with R `lm()`.  Interpret the coefficients in context.

```{r, class.source="rsoln", class.output="rsoln"}
r <- cor(x, y)
xbar <- mean(x); ybar <- mean(y)
s_x <- sd(x); s_y <- sd(y)

est_slope <- r * (s_y / s_x)
est_intercept <- ybar - (xbar * est_slope)

c(est_slope, est_intercept)
```

[The linear model is $\hat{y}_i = 3.4 + 2.22(x_i)$.  3.4 is the therotical happiness score for a country with 0 GDP per capital, and 2.22 is the expected increase in happiness score for 1 unit increase in GDP per capita.  We get the same results with `lm()`.]{.soln}

```{r, class.source="rsoln", class.output="rsoln"}
happiness_mod <- lm(Score ~ GDP.per.capita, data = happiness)

happiness_mod
```

(c) Using your linear model, predict the happiness index for theoretical countries with 0.75, 1.25, and 2 GDP per capita.  Which of these predictions would you consider to be the least reliable, and why?

```{r, class.source="rsoln", class.output="rsoln"}
gdp <- c(0.75, 1.25, 2)

# Using predict function
predict(happiness_mod, newdata = tibble(GDP.per.capita = gdp))

# By hand
3.4 + (2.22 * gdp)
```

[The predictions for happiness index are approximately 5.065, 6.175, and 7.84.  I am most skeptical of the prediction at x = 2 GDP per capita, because that value is well outside the range of our original data.  We have only verified that the linear model is correct for values of x within [0, 1.684].  Predicting at a value of 2 is extrapolation.]{.soln}

(d) What are the conclusions we can make about the relationship between GDP per capita and happiness index, based on the linear model?  Does money buy happiness?

[Based on our linear model, we can say that countries with higher GDP per capita generally tend to have a higher happiness index.  However, we cannot make any conclusions about causality between the two variables.  It is possible that high GDP and high happiness are both related to other aspects of the country (e.g. more stable government and more social support).]{.soln}



