---
title: "Homework 10"
author: "Student's Name Here"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE,
                      fig.height = 3)
library(tidyverse)
```

```{r, echo=F}
old <- theme_get()

theme_set(theme(panel.background = element_rect(fill = "#ededfb")))
```


# Preliminaries

- This file should be in `STAT240/homework/hw10` on your local computer.
- Download `parental_leave.csv` and `happiness_2019` to `STAT240/data`.


# Problem 1

Two sets of data were used in the equal-variances T hypothesis test of

$$H_0: \mu_1 - \mu_2 = 2 \quad \text{versus} \quad H_A: \mu_1 - \mu_2 < 2$$

We know that $s_1 = 2, s_2 = 3, n_1 = 16$, and $n_2 = 15$. If the p-value is 0.001298716, what is the observed difference in means, $\bar{x}_1 - \bar{x}_2$?


# Problem 2

Continue working with the parental leave data from homework 9.

```{r}
leave <- read_csv("../../data/parental_leave.csv")

# Remove spaces from column names
names(leave) <- make.names(names(leave))

# Tidy numeric columns
leave <- leave %>%
  mutate_at(vars(Paid.Maternity.Leave, Unpaid.Maternity.Leave,
                 Paid.Paternity.Leave, Unpaid.Paternity.Leave),
            as.numeric) %>%
  replace(is.na(.), 0)
```

(a) Re-do the basic data tidying from homework 9 problem 5(a): Mutate a column for total paid leave by adding the weeks of paid maternity and paternity leave. Optionally, give the columns shorter names that are easier to reference.

(b) Consider comparing companies in "Educational Services: College & Universities" (ESCU) to companies in "Healthcare: Hospitals & Clinics" (HHC).  Create and print a table summarizing the mean, standard deviation, and sample size for both groups.

(c) Assess the equal variances assumption for these two groups of data by comparing the ratio of standard deviations.  Even if the standard deviations are similar "enough", why might we want to not assume equal variances?


# Problem 3

Assuming unequal variances, build a 95% CI for the difference in average total weeks of paid leave between the ESCU and HHC industries and interpret the results. You must do this by hand, but you can check your results with `t.test()`.


# Problem 4

Does the educational industry give more paid leave than the healthcare industry?  Perform a hypothesis test of 
$$H_0: \mu_{ESCU} - \mu_{HHC} = 0 \quad \text{versus}\quad H_A: \mu_{ESCU} - \mu_{HHC} > 0$$
with $\alpha = 0.05$, assuming unequal variances.  You must do this by hand, but you can check your results with `t.test()`.


# Problem 5

Repeat the by-hand test in problem 4, but instead write hypotheses for $\mu_{HHC} - \mu_{ESCU}$ (the order of subtraction is switched).  Show that you get identical results to problem 4.


# Problem 6

Now, consider the "Technology: Software" industry.  Perform the appropriate hypothesis test to determine whether there is a difference in weeks of paid maternity and paternity leave within this industry.  You must do this by hand, but you can check your results with `t.test()`.


# Problem 7

Match the six values of correlation to the scatterplots in `p7_choices.png`. Briefly justify your choices.

- $r = -0.85$

- $r = -0.74$

- $r = 0.08$

- $r = 0.44$

- $r = 0.98$

- $r = 1$


# Problem 8

The [Happiness index](https://www.kaggle.com/datasets/sougatapramanick/happiness-index-2018-2019) data contains information on the happiness index of different countries from 2018-2019.  We will focus on the happiness index (`Score`) and GDP per capita of countries in 2019.

```{r}
happiness <- read_csv("../../data/happiness_2019.csv")

# Remove spaces from column names
names(happiness) <- make.names(names(happiness))
```

(a) Create a scatterplot of GDP per capita (x) versus happiness index (y) and calculate the correlation between length and weight.  Comment on the strength and magnitude of the linear relationship, and whether a linear model would be appropriate for this data.

(b) Calculate the slope and intercept of a least-squares linear regression model for GDP per capita versus happiness index.  Do this with the "by hand formulas" and check your work with R `lm()`.  Interpret the coefficients in context.

(c) Using your linear model, predict the happiness index for theoretical countries with 0.75, 1.25, and 2 GDP per capita.  Which of these predictions would you consider to be the least reliable, and why?

(d) What are the conclusions we can make about the relationship between GDP per capita and happiness index, based on the linear model?  Does money buy happiness?


