---
title: "Discussion 11"
author: "Solutions"
output: html_document
---

```{css, echo = F}
.rsoln {
background-color: #E6E6FA;
}

.soln {
color: #5858d0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE,
                      error = TRUE)

library(tidyverse)
```

```{r, echo=F}
old <- theme_get()

theme_set(theme(panel.background = element_rect(fill = "#ededfb")))
```

# Preliminaries

- This file should be in `STAT240/discussion/ds11`.
- This assignment should be completed in your discussion groups, which can be found on the "People" tab on Canvas.

The goal of today's discussion will be to think about different ways of studying two separate populations.  You'll also get to see a few highly questionable linear models.


# Problem 1
#### (approx. 20 minutes)

In statistics, we often have an idea of what a desirable or informative result looks like. We ask, "What qualities must our analysis have to acheive this result?" Perform these calculations for the three separate examples below.

(a) Consider building a T confidence interval with $n = 35$ observations based on a sample with sample sd $s = 8.4$. What is the biggest confidence level (smallest $\alpha$) we can use if we want to have margin of error less than 3?

<span style="color:#5858d0"> The margin of error is given by the critical value $t_{35, \alpha/2}$ multiplied by the standard error $\frac{s}{\sqrt{n}}$.  We have
$$MoE \;=\; 3 \;=\; t\cdot \frac{8.4}{35} \quad \Rightarrow \quad t \;=\; 2.113$$
So, 2.113 is the quantile that cuts off $\alpha/2$ area in the upper tail of the T curve. To find $\alpha/2$, use `pt()`. </span>

```{r, class.source="soln", class.output="soln"}
pt(2.113, df = 34, lower.tail = F)
```

<span style="color:#5858d0"> We get $\alpha/2 = 0.02$ which means $\alpha = 0.04$. This corresponds to 96% confidence.  We need a CI with level 96% or lower to get our desired margin of error. </span>

(b) A 95% Wald confidence interval will be used to obtain an estimate for an unknown population proportion $p$. What is the smallest sample size that will guarantee a margin of error of at most 0.08?

<span style="color:#5858d0"> The margin of error for a 95% Wald proportion Z CI is given by
$$1.96\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$
where 1.96 is the 97.5 percentile of the standard normal.  We set this equal to 0.08 and solve for $n$. The last value we need is $\hat{p}$. We want to guarantee that our margin of error will be 0.08 or less, for any $\hat{p}$. We have to pick the "worst" $\hat{p}$ that corresponds to the largest possible standard error, to make sure our $n$ will always work. The standard error is largest when $\hat{p} = 0.5$.  So,
$$0.08 = 1.96\cdot \sqrt{\frac{0.5(0.5)}{n}} \quad \Rightarrow \quad n = \Big(\frac{1.96}{0.08}\Big)^2 \cdot 0.25 = 150.06$$ </span>

```{r, class.source="soln", class.output="soln"}
n <- (1.96/0.08)^2 * 0.25
n
```

<span style="color:#5858d0"> We can then round this up to be $n = 151$. Because this value of $n$ gives us a small enough margin of error for the worst-case $\hat{p}$, the margin of error will also be smaller than 0.08 for any other value of $\hat{p}$. </span>

(c) We are performing a one-sided T hypothesis test of 
$$H_0: \mu = 5 \quad \text{versus}\quad H_A: \mu > 5$$
with $s = 43.1$ and $n = 150$. How large does the observed sample mean $\bar{x}$ need to be to see a one-sided p-value of 0.05?  *Start with the formula for a test statistic, then solve for $\bar{x}$.  What value goes on the other side of your equation?*

<span style="color:#5858d0"> For $n = 150$, the p-value of 0.05 is calculated on a $t_{149}$ distribution. The test statistic must correspond to the 95th percentile of $t_149$ for the p-value (area above) to be 0.05. </span>

```{r, class.source="soln", class.output="soln"}
test_stat <- qt(0.95, df = 149)
test_stat
```

<span style="color:#5858d0"> So, a test statistic of 1.655 or greater results in us rejecting the null hypothesis. Now, we can use the test statistic formula with 5, $s$, and $n$ and solve for $\bar{x}$. 
$$1.655 = \frac{\bar{x} - 5}{43.1/\sqrt{150}} \quad \Rightarrow \quad \bar{x} = 1.655\cdot \frac{43.1}{\sqrt{150}} + 5 = 10.825$$ </span>

```{r, class.source="soln", class.output="soln"}
xbar <- test_stat * (43.1/sqrt(150)) + 5
xbar
```

<span style="color:#5858d0"> To get a p-value greater than 0.05, we would need to observe a sample mean $\bar{x}$ of at least 10.825, or 5.825 higher than the
null value of 5. </span>

(d) In the previous three parts, we worked with the significance level $\alpha$, the sample size $n$, and the sample mean $\bar{X}$. Which of these three quantities would it be most logical to manipulate, as a researcher? Explain your choice.

<span style="color:#5858d0"> It would make the most sense to adjust the sample size $n$.  We can improve our estimates and our hypothesis tests by taking a larger sample of data.  It does not make sense to manipulate $\bar{X}$, since that value comes from observing data. </span>

<span style="color:#5858d0"> Even though we can control $\alpha$, it's also not a good idea to change this quantity to get a better result.  Best practice is to set $\alpha$ ahead of time according to how much risk we are comfortable with. </span>


# Problem 2
#### (approx. 15 minutes)

Consider exploring the formula for Welch degrees of freedom for a two-sample T test.

(a) Create a dataframe with the following columns: 

- `s1`, with values 1 through 40, repeated 40 times.
- `s2`, with values 1 through 40, repeated 40 times, so that we have 1600 unique combinations of `s1` and `s2`.
- `w_df`, the Welch degrees of freedom evaluated for each combination of `s1` and `s2`.  Use n1 = n2 = 20 and `round()` to 3 decimal points to avoid precision errors with R.

    Think about how to use the `rep()` function cleverly to get every different combination of `s1` and `s2` without any repeat rows.

```{r, class.source="rsoln", class.output="rsoln"}
welch_df_data <- tibble(
  s1 = rep(1:40, 40),
  s2 = rep(1:40, each = 40),
  
  w_numerator = (s1^2/20 + s2^2/20)^2,
  w_denominator = (s1^4/(20^2*(20-1)) + s2^4/(20^2*(20-1))),
  
  w_df = round(w_numerator / w_denominator, 3)
)
```

(b) What are the minimum and maximum values of `w_df` in your the data?  What are the corresponding values of `s1` and `s2`?

```{r, class.source="rsoln", class.output="rsoln"}
welch_df_data %>%
  slice_min(w_df)

welch_df_data %>%
  slice_max(w_df)
```

[The smallest value of df, about 19, occurs at the biggest difference between `s1` and `s2`.  The maximum value of the df, which is n1 + n2 - 2 = 38, occurs in every instance where s1 and s2 are equal.]{.soln}

(c) Use ggplot with `geom_tile()` to create a heat-map plot of your dataframe with `s1` and `s2` on the x and y axes and the tiles colored by `w_df` in a color scheme of your choice. 

```{r, class.source="rsoln", class.output="rsoln"}
ggplot(welch_df_data, aes(x = s1, y = s2, fill = w_df)) +
  geom_tile() +
  scale_fill_viridis_c()
```

(d) Based on the results of (b) and (c), interpret the relationship between the size of `s1` and `s2` and the Welch degrees of freedom.

[There is a bigger "penalty" to the Welch df when the two sample standard deviations are more different.  We see that the Welch df are always the same when the two sample SDs are equal, regardless of the size of the SD. (In a T test that assumes equal population variances, the degrees of freedom are simply n1 + n2 - 2).]{.soln}

[In the plot, we see that the bigger values for df "fan out" as we move to the top right.  So, the amount that the degrees of freedom are penalized depends on the magnitude of s1 and s2 - we seem to care more about the relative difference between the values rather than just the absolute difference.]{.soln}


# Problem 3
#### (approx. 10 minutes)

Identify whether each of the datasets described below are paired or independent samples.  Briefly explain your choices.

(a) Measurements of annual income taken both before and after a two-year training course for a random sample of 100 people who took the course.

[Paired - the samples are not independent because we are measuring the same person's income before and after the course.  The analysis should be done at the level of each person.]{.soln}

(b) Measurements of annual income for a random sample of 100 people who took the two-year training course and another random sample of 100 people who did not take the course.

[Independent - in this case we are studying completely different people in the group who took the course and the group who did not, so there is no meaningful way to pair the data.]{.soln}

(c) Measurements of annual income for both individuals in pairs formed by assigning 100 people to pairs at random.

[Independent - just like part (b), there is no special meaning to this pairing that we expect to affect our analysis, so we should consider the two groups as independent.]{.soln}

(d) Measurements of annual income 100 randomly selected pairs of twins that have one female and one male twin.

[Paired - we cannot assume that siblings are independent of each other since they probably share a lot of background and upbringing that might affect their financial situation. We should do our analysis on the level of each pair of twins.]{.soln}

(e) Measurements of annual income recorded for both partners of 100 randomly selected couples.

[Paired - the groups are not independent because the individuals have chosen to enter a relationship with an individual from the other group. We should do the analysis on the level of each couple since the partners might have a similar job or share financial habits.]{.soln}


