---
title: "live: Inference on means"
author: "Sahifa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


What is the true average time that it takes a participants of the Boston Marathon in the age range of "18-34" to finish the marathon?


```{r}
bm

bm_summary = bm %>% 
  filter(age_range == "18-34", Year == 2010) %>% 
  summarize(averageTime = mean(total_time), sdTime = sd(total_time), n = n())

xbar = 223.8312
s = 40.90398
n = 6448

# 95% CI for the true average time taken to complete BM.

c = 0.95
cv = qt(c+(1-c)/2, n-1) # for this example qnorm(c+(1-c)/2) is almost the same as the cv we have here because our sample size is huge
cv

se = s/sqrt(n)
se

xbar + c(-1, 1)*cv*se
```

Is the true average time taken to finish the Boston Marathon equal to 4 hrs (240 mins)?

$$H_0: \mu = 240 \\
H_{\alpha}: \mu \neq 240$$

```{r}
mu_h0 = 240

t = (xbar-mu_h0)/(s/sqrt(n))
t
p_val = 2*pt(abs(t), n-1, lower.tail = F)
p_val # alpha = 0.05, we reject H_0
```

```{r}
times = bm %>% filter(age_range == "18-34" & Year == 2010) %>% pull(total_time)

t.test(times) # H_0: mu = 0

t.test(times, mu = 240, alternative = "less", conf.level = 0.9) # mu is for changing the val for hypothesis test, alternative is for changing the alternative hypo (by default it is two sided), conf.level = to change the confidence level of confidence interval
```
# Inference on differnce of means

# Unequal variances

$X$ = time taken by male participants in the age range of "18-34" of BM $\sim D_x(\mu_x, \sigma_x)$
$Y$ = time taken by female participants in the age range of "18-34" of BM $\sim D_y(\mu_y, \sigma_y)$

$\mu_x-\mu_y$ most likely negative

```{r}
bm_twomeans_summary = bm %>% 
  filter(age_range == "18-34") %>% 
  group_by(sex) %>% 
  summarize(avgTime = mean(total_time), sdTime = sd(total_time), n = n())

bm_twomeans_summary
```

```{r}
ybar = bm_twomeans_summary$avgTime[1]
xbar = bm_twomeans_summary$avgTime[2]
sy = bm_twomeans_summary$sdTime[1]
sx = bm_twomeans_summary$sdTime[2]
ny = bm_twomeans_summary$n[1]
nx = bm_twomeans_summary$n[2]
```

```{r}
# 95% CI for the difference of means: pe +_ cv*se

p.e = xbar - ybar
se = sqrt(sx^2/nx + sy^2/ny)
w = (sx^2/nx + sy^2/ny)^2 / (sx^4/(nx^2*(nx-1)) + sy^4/(ny^2*(ny-1)))
cv = qt(0.95+(1-0.95)/2, w)
cv

p.e + c(-1, 1)*cv*se
```

- Hypothesis test

$$H_0: \mu_x-\mu_y = 0 \\
H_{\alpha}: \mu_x - \mu_y < 0$$


```{r}
t = (p.e - 0)/se
t

p_val = pt(t, w)
p_val # alpha = 0.05, p-val < alpha, we reject H_0
```


```{r}
times_x = bm %>% 
  filter(age_range == "18-34", sex == "male") %>% select(total_time)

times_y = bm %>% 
  filter(age_range == "18-34", sex == "female") %>% select(total_time)


t.test(times_x, times_y, alternative = "less")
```
# equal variance

```{r}
times_x = bm %>% 
  filter(age_range == "18-34", sex == "female", Year == "2010") %>% select(total_time)

times_y = bm %>% 
  filter(age_range == "18-34", sex == "female", Year == "2011") %>% select(total_time)

t.test(times_x, times_y, var.equal = T)
t.test(times_x, times_y, var.equal = T, alternative = "greater")
```

```{r}
thirteen = c(44.1, 59.0, 65.9, 58.7, 49.3)
fourteen = c(46.3, 60.5, 68.2, 59.4, 50.6)

diff = thirteen - fourteen

t.test(diff)
t.test(thirteen, fourteen, paired = T, conf.level = 0.9)
```








