---
title: "live: Inference on means"
author: "Sahifa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 004

# Infernce on single mean

What is the true average time that it takes a participants of the Boston Marathon in the age range of "18-34" to finish the marathon? 

```{r}
bm

bm_summary = bm %>% filter(age_range == "18-34", Year == 2010) %>% summarise(sample_mean = mean(total_time), sample_sd = sd(total_time), n = n())

xbar = 223.8312
s = 40.90398
n = 6448

# 95% CI fo rthe true avg time for the BM

c = 0.95
cv = qt(c+(1-c)/2, n-1) # since our sample size is big, the quantile of t distributioj is going to be same as the normal distribution

cv

se = s/sqrt(n)

xbar + c(-1, 1)*cv*se
```


Hypothesis test: Is the true average time it takes a participants of the Boston Marathon in the age range of "18-34" to finish the marathon 4 hrs?


$$H_0: \mu = 240 \\
H_{alpha}: \mu \neq 240$$


```{r}
t = (xbar - 240)/se
t

2*pt(abs(t), n-1, lower.tail = F)
```


```{r}
times = bm %>% filter(age_range == "18-34", Year == 2010) %>% select(total_time)

t.test(times)

t.test(times, mu = 240, alternative = "less", conf.level = 0.9)

t.test(times, mu = 224, alternative = "less")
```
# Inference on differnece of means

# unequal variances
$\sigma_x \neq \sigma_y$

$\mu_x - \mu_y$ mostly be a negative number

```{r}
times_male = bm %>% filter(age_range == "18-34", sex == "male") %>% select(total_time)

times_female = bm %>% filter(age_range == "18-34", sex == "female") %>% select(total_time)

bm_summary_two_means = bm %>% filter(age_range == "18-34") %>% group_by(sex) %>% summarise(sample_avg = mean(total_time), sample_s = sd(total_time), n = n())

bm_summary_two_means
```

```{r}
ybar = bm_summary_two_means$sample_avg[1]
xbar = bm_summary_two_means$sample_avg[2]
sy = bm_summary_two_means$sample_s[1]
sx = bm_summary_two_means$sample_s[2]
ny = bm_summary_two_means$n[1]
nx = bm_summary_two_means$n[2]
```

```{r}
# 95% CI 

pe = xbar - ybar
se = sqrt(sx^2/nx + sy^2/ny)
c = 0.95
w = (sx^2/nx + sy^2/ny)^2 / (sx^4/(nx^2*(nx-1)) + sy^4/(ny^2*(ny-1)))
cv = qt(c+(1-c)/2, w)
cv
pe + c(-1, 1)*cv*se
```
- Hypothesis testing

$$H_0: \mu_x - \mu_y = 0 \text{ vs } \mu_x=\mu_y \\
H_{\alpha}: \mu_x - \mu_y < 0 \text{ vs } \mu_x<\mu_y$$
```{r}
t = (pe - 0)/se # has a t distribution or that your null dist is t
t

p_val = pt(t, w)
p_val # alpha = 0.05, p-val< 0.05, we reject H_0
```

```{r}
t.test(times_male, times_female, alternative = "less", conf.level = 0.9)
```

# equal variance

```{r}
times_2010 = bm %>% filter(age_range == "18-34", sex == "female", Year == "2010") %>% select(total_time)
  
times_2011 = bm %>% filter(age_range == "18-34", sex == "female", Year == "2011") %>% select(total_time)

t.test(times_2010, times_2011)

t.test(times_2010, times_2011, var.equal = T)
```

# Paired t test

```{r}
thirteen = c(44.1, 59.0, 65.9, 58.7, 49.3)
fourteen = c(46.3, 60.5, 68.2, 59.4, 50.6)

diff = thirteen-fourteen
diff

t.test(diff)
t.test(thirteen, fourteen, paired = T)
```























