---
title: "live_dplyr_week3"
author: "Sahifa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)

# install.packages("magrittr")
library(magrittr)
```
# Monday

# Pipe operator: %>% 

MAC: cmd + shift + M
Windows: ctrl + shift + M

```{r}
3*5^2 + 2

# magrittr has functions for all basic mathematical operations

add(multiply_by(raise_to_power(5, 2), 3), 2)

5 %>% raise_to_power(2) %>% multiply_by(3) %>% add(2)
```


# datasets

```{r}
grocery_prices = read_csv("../data/grocery-prices.csv")

grocery_list = read_csv("../data/grocery-list.csv")
```

# Row functions:

```{r}
grocery_prices
```


```{r}
grocery_prices %>% arrange(price) # this arranges the price column in the increasing order

grocery_prices %>% arrange(item)
```
```{r}
grocery_prices %>% arrange(desc(price))
grocery_prices %>% arrange(-price)
```

- filter():

```{r}
grocery_prices %>% filter(price > 1)
```

# Wednesday

# Column Operations: 

- mutate(): this functions creates columns in your dataset, format is similar to tibble

```{r}
grocery_prices %>% mutate(price_after_tax = price*0.02, total_price = price+price_after_tax) 
```

- select(): this function can create subsets of your dataset (not keep all the columns in the data)

```{r}
grocery_prices %>% 
  mutate(price_after_tax = price*0.02, total_price = price+price_after_tax) %>% 
  select(item, type, total_price)

grocery_prices %>% 
  mutate(price_after_tax = price*0.02, total_price = price+price_after_tax) %>% 
  select(- price, - price_after_tax)

grocery_prices %>% 
  mutate(price_after_tax = price*0.02, total_price = price+price_after_tax) %>% 
  select(- c(price, price_after_tax))

grocery_prices %>% 
  mutate(price_after_tax = price*0.02, total_price = price+price_after_tax) %>% 
  select(- c(price:price_after_tax))

grocery_prices %>% 
  mutate(price_after_tax = price*0.02, total_price = price+price_after_tax) %>% 
  select(price:price_after_tax)
```

- relocate(): this function relocates columns and preserves the order that you provide the columns in 

```{r}
grocery_prices %>% 
  mutate(price_after_tax = price*0.02, total_price = price+price_after_tax) %>% 
  relocate(item, type, total_price)

grocery_prices %>% 
  mutate(price_after_tax = price*0.02, total_price = price+price_after_tax) %>% 
  relocate(type, total_price, .after = item)

grocery_prices %>% 
  mutate(price_after_tax = price*0.02, total_price = price+price_after_tax) %>% 
  relocate(type, total_price, .before = price)
```

- rename(): this function renames columns, format is similar tibble, new_col_name = old_col_name

```{r}
grocery_new = grocery_prices %>% 
  mutate(price_after_tax = price*0.02, total_price = price+price_after_tax) %>% 
  rename(tax = price_after_tax)

grocery_new
```

# Operations on rows

- arrange():

```{r}
grocery_new %>% arrange(total_price)

grocery_new %>% arrange(- total_price)

grocery_new %>% arrange(desc(total_price))
```

- filter(): this function filtering rows that satisfy the condition/logical statement 

```{r}
grocery_new %>% filter(price>=1)

# extract rows that have the type of item as vegetable
grocery_new %>% filter(type == "vegetable")

grocery_new %>% filter(!is.na(price)) # all the rows that don't have NA in price
```

```{r}
starwars %>% filter(!is.na(hair_color) & !is.na(mass))
```

- slice_max() and slice_min(): these functions extract 'n' rows that have the min or max in the specified column

```{r}
grocery_new %>% slice_max(total_price, n = 3)
```

- Operations on groups:



```{r}
view(grocery_new)
```


- summarise(): this function will create a summary table

```{r}
grocery_new %>% summarise(avg_price = mean(price), sd_price = sd(price), median_price = median(price), max_total = max(total_price)) # summary functions are function sthat create one output for the whole vector

grocery_new %>% summarise(avg_price = mean(price), sd_price = sd(price), median_price = median(price), max_total = max(total_price), sqrt = sqrt(price))
```



