---
title: "Section 18: Chess Case Study"
author: "Solutions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      error = TRUE, fig.height = 4)
library(tidyverse)

theme_set(theme_minimal())
```


```{r}
games <- read_csv("../../data/chess_tidy.csv")
```


# 1. Proportion Inference

## 1.1 Which player wins more often?

```{r}
games %>%
  filter(winner != "draw") %>%
  count(winner)
```

$$H_0: p_w = 0.5 \quad \text{versus} H_A: p_w > 0.5$$

```{r}
# Summary values + observed proportion
x <- 6302
n <- 5841 + 6302
phat <- x / n

# Z test statistic
test_stat <- (phat - 0.5) / sqrt(0.5*0.5 / n)

# One-sided p-value
pnorm(test_stat, lower.tail = F)
```

We have evidence at the 1% level that white wins more than half the time (i.e. $p_w > 0.5$).


## 1.2 How does victory type affect the winner?

```{r}
games %>%
  filter(victory_status %in% c("mate", "resign")) %>%
  count(victory_status, winner)
```

For a Z confidence interval to be valid, both groups must have a large number of successes and failures.  This seems to be well-met in this case.  Let group 1 be "mate" and group 2 be "resign".

```{r}
# Summary values
x1 <- 2192; n1 <- 1898 + 2192
x2 <- 3640; n2 <- 3430 + 3640

phat1 <- x1/n1; phat2 <- x2/n2

pt_est <- phat1 - phat2

# Standard error for a difference in proportions
se <- sqrt(phat1*(1-phat1)/n1 + phat2*(1-phat2)/n2)

# Standard normal critical value for 99% confidence
cv <- qnorm(0.995)

c(pt_est - cv*se, pt_est + cv*se)
```

We are 99% confidence the true difference in white's winrate is between (-0.004, 0.046).  The interval contains 0, so there does not appear to be a significant difference in white's advantage in games that end in mate vs those that end in resignation.


# 2. Means Inference

## 2.1 Average number of turns

```{r}
games %>%
  ggplot(aes(x = turns)) +
  geom_histogram()

turn_summary <- games %>%
  summarize(xbar = mean(turns),
            s = sd(turns), 
            n = n())

turn_summary
```

The histogram is right-skewed, but it is still acceptable to do a T test because the CLT sayt that the sample mean Xbar will be approximately normal.

```{r}
# Get summary values from table
pt_est <- turn_summary$xbar
s <- turn_summary$s
n <- turn_summary$n

# Estimated standard error
se <- s / sqrt(n)

# T critical value for 99% confidence
cv <- qt(0.995, df = n-1)

c(pt_est - cv*se, pt_est + cv*se)

t.test(games$turns, conf.level = 0.99)
```

We are 99% confident that the true average number of turns in a chess game is between (61.65, 63.18).  This is confirmed in `t.test()`.


## 2.2 Average number of turns based on starting time

Let $\mu_1$ be the average number of turns taken for games with 15 minutes or less of starting time and $\mu_2$ be the average number of turns taken for games with over 15 minutes of starting time.  We want to test

$$H_0: \mu_1 - \mu_2 = 0 \quad \text{versus} \quad H_A: \mu_1 - \mu_2 < 0$$


```{r}
# First, mutate a variable for starting time "type"

games <- games %>%
  mutate(starting_type = case_when(starting_minutes <= 15 ~ "short",
                                   starting_minutes > 15 ~ "long"))

# Calculate summary statistics
turn_summary <- games %>%
  group_by(starting_type) %>%
  summarize(xbar = mean(turns),
            s = sd(turns), 
            n = n()) %>%
  arrange(desc(starting_type))

turn_summary
```

```{r}
# Get summaries from table
xbar1 <- turn_summary$xbar[1]
xbar2 <- turn_summary$xbar[2]
s1 <- turn_summary$s[1]
s2 <- turn_summary$s[2]
n1 <- turn_summary$n[1]
n2 <- turn_summary$n[2]

# Compare s1 and s2
s1 / s2
```

We should do a two independent sample T test.  We can additionally decide if we want to make the equal variance assumption.  Since s1 and s2 are very similar, we can proceed with the equal variances T test.  But a Welch T test would also be valid.

```{r}
# Calculate pooled sd
sp <- sqrt(((n1 - 1)*s1^2 + (n2 - 1)*s2^2) / (n1 + n2 - 2))

se <- sp * sqrt(1/n1 + 1/n2)

# Calculate T test statistic
test_stat <- (xbar1 - xbar2) / se

# Calculate one-sided p-value
pt(test_stat, df = n1 + n2 - 2)
```

The observed effect in our data is in the opposite direction of our hypotheses: short games tend to have more turns than long games.  So, we end up with a very large, very non-significant p-value.


## 2.3 Player ratings

We should perform a paired t-test to look at the difference in rating between each game.

```{r}
# Find rating difference (white-black)
games <- games %>%
  mutate(rating_diff = white_rating - black_rating)

rating_summary <- games %>%
  summarize(xbar = mean(rating_diff),
            s = sd(rating_diff),
            n = n())

rating_summary
```

```{r}
# Get summary values from table
xbar <- rating_summary$xbar
s <- rating_summary$s
n <- rating_summary$n

# Estimated standard error
se <- s / sqrt(n)

# Test statistic
test_stat <- (xbar - 0) / se

# Two-sided p-value
2*pt(test_stat, df = n-1, lower.tail = F)

# Check with t.test
t.test(games$rating_diff)
```

We get a non-significant p-value of 0.814.  So we have no evidence that black or white is systematically assigned to the player with the better rating.


# 3. Linear Regression

```{r}
chess_mod <- lm(turns ~ abs_rating_diff, data = games)

summary(chess_mod)
```

The base number of turns between two evenly matched players is 65.75.  For a one-point rating difference between the players, we would expect the number of turns to decrease by 0.022.


# 3.1 Slope confidence interval

```{r}
# Get slope and slope SE from summary output
pt_est <- summary(chess_mod)$coefficients[2, 1]
se <- summary(chess_mod)$coefficients[2, 2]

# Find 99% T critical value
n <- nrow(games)
cv <- qt(0.995, df = n-2)

c(pt_est - cv*se, pt_est + cv*se)
```

We are 99% confident that the true slope for rating difference and number of turns is within (-0.027, -0.017).  The confidence interval is entirely negative.


# 3.2 Validate the linear model

```{r}
games <- games %>%
  mutate(resids = resid(chess_mod))

ggplot(games, aes(x = abs_rating_diff, y = resids)) +
  geom_point() +
  geom_hline(yintercept = 0)
```


The residual plot appears to have some issues with non-normality but especially with non constant variance. The residuals are much more spread out for pairings with a small rating difference. Here are some possibilities that might help:

- Transform the data
- Filter to a smaller selection of games that come from the same "population"
- Find a different model



