---
title: "Discussion 12"
author: "Group Member Names Here"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE,
                      error = TRUE)

library(tidyverse)
```

# Preliminaries

- This file should be in `STAT240/discussion/ds12`.
- The file `example_lm_data.csv` should be in `STAT240/data`.
- This assignment should be completed in your discussion groups, which can be found on the "People" tab on Canvas.

The goal of today's discussion will be to build a stronger understanding of the meaning of correlation in a linear model and how to use the model for different types of predictions.


# Problem 1
#### (approx. 15 minutes)

The dataset `example_lm_data.csv` contains four sets of (x, y) pairs.  
```{r}
lm_data <- read_csv("../../data/example_lm_data.csv")
```

(a) Use R to fit a linear model on each dataset (four linear models total).  Also calculate the correlation between each set of (x, y) pairs.  What do you notice about the four sets of data?

(b) Now, use `ggplot()` with `geom_point()` and `geom_smooth()` to make scatterplots of each dataset with its respective linear model.  What can be said about the data now that we can see it?  Which set of data is most appropriately described by the linear model?

After completing the activity, you can read more about this example [here](https://medium.com/@marc.bolle/datasaurus-dozen-visualization-using-python-d328dad64d20).


# Problem 2
#### (approx. 20 minutes)

The code below (hidden in the .html) creates a ficitonal dataset of 35 students and their midterm and final exam scores (out of 100) and creates a basic scatterplot of the data.

```{r, include=FALSE}
exam_scores <- tibble(
  midterm = c(85.4, 91.0, 83.9, 83.9, 84.0, 85.8, 76.2, 83.9, 73.5,
              80.0, 83.9, 85.0, 65.0, 79.4, 91.5, 90.4, 79.4, 80.7,
              76.0, 72.3, 87.3, 82.1, 88.2, 69.2, 72.9, 68.9, 86.2,
              90.8, 83.7, 72.4, 79.7, 82.7, 80.6, 94.5, 97.5),
  final = c(82.9, 95.8, 79.6, 89.2, 81.2, 86.9, 69.1, 86.1, 76.2,
            83.2, 81.4, 83.3, 65.3, 69.4, 83.8, 89.2, 83.6, 80.8,
            77.3, 71.9, 85.7, 89.4, 85.3, 74.0, 73.8, 73.9, 88.5,
            93.5, 85.6, 76.8, 69.8, 73.7, 80.4, 98.4, 88.2)
)

ggplot(exam_scores, aes(x = midterm, y = final)) + 
  geom_point()
```


(a) Using R, calculate the mean and standard deviation of both the midterm and final exam scores, and calculate the correlation between the two scores.

(b) Is it true that students had above-average midterm scores also had above-average final exam scores?  Is it true that students who had above-average midterm scores showed an improved score on the final?

(c) Using the basic "by-hand" formulas (i.e. not using R `lm()`), calculate a linear model to predict final exam score (y) in terms of midterm score (x).

(d) Suppose a student scored *one standard deviation above the class average* for their midterm score.  What does the model predict their final exam score to be?  What is this student's final exam score in terms of the average and standard deviation final exam score for the class?

(e) For linear models in general, if a value $x_i$ is $z_i$ standard deviations above the average of X, the predicted value $\hat{y}_i$ will be how many standard deviations above the average of Y?

(f) In a famous anecdote, flight instructors claim that punishment and shame is the most effective teaching tool.  Student pilots that were praised for good performance tended to do worse on their next flight, and student pilots that were berated for poor performance tended to do better on their next flight.  How can we explain this phenomenon with regression?


---

# Final Exam: Thursday, May 7

Use the remaining time in discussion to work on writing your cheat sheet for the final exam.  You can bring *two* standard sheet of paper with anything you want written on both sides, and you are welcome to work with others to decide what you want to put on your sheet.

The final exam will be cumulative, so topics from any point in the semester are eligible to appear on the exam.  However, there will be a focus on chapters covered after the second midterm (Miranda's sections 11-16).

